---
Title: EDA on Prosper Loans
---

## The Data Set {#top}

This report explores loan data provided by the online lender Prosper.The data 
spans from November, 2005 to March, 2014, with 113,937 observations. 

The data used in this exploration was tidied and provided by Udacity for the EDA
learning section of their online Data Analysis degree. The original set included
81 variables, and my instructions were to evaluate 10-15 variables for my EDA. I 
chose the following variables: 

* Loan amount
* Loan term months
* Loan status current
* Employment status
* Credit score lower
* Credit score upper
* Debt income ratio
* Stated income monthly
* Principal paid
* Interest and fees paid
* Estimated return

I follow a standard EDA process in this project, progressing from 
[univariate](#univariate) to [bivariate](#bivariate) to 
[multivariate](#multivariate) explorations. In the [summary](#summary) section, 
I pick three visualizations from the EDA. These graphs are what I would 
communicate to stakeholders to share what I learned during my exploration.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(tidyverse)
library(GGally)
```

```{r include=FALSE}
# Load the Data
loans_full_set <- read.csv('prosperLoanData.csv')

# Select variables of interest(instructions are to look at 10-15 variables)
loans <- select(loans_full_set, 
                LoanOriginalAmount, Term, LoanStatus, EmploymentStatus, 
                EmploymentStatusDuration,  CreditScoreRangeLower, 
                CreditScoreRangeUpper, DebtToIncomeRatio, StatedMonthlyIncome,  
                LP_CustomerPrincipalPayments, LP_InterestandFees, 
                EstimatedReturn)

loans <- rename(loans,
       loan_amount = LoanOriginalAmount,
       loan_term_months = Term,
       loan_status_current = LoanStatus,
       employment_status = EmploymentStatus,
       employment_duration_months = EmploymentStatusDuration,
       credit_score_lower = CreditScoreRangeLower,
       credit_score_upper = CreditScoreRangeUpper,
       debt_income_ratio = DebtToIncomeRatio,
       stated_income_monthly = StatedMonthlyIncome,
       principal_paid = LP_CustomerPrincipalPayments,
       interest_and_fees_paid = LP_InterestandFees,
       estimated_return = EstimatedReturn
       )

```

# Univariate Exploration {#univariate}

[Back to Top](#top)

Loan amounts cluster vary, with many loans equal to or less than $10,000. The 
most frequent loan amounts are $4,000, $15,000, $10,000, and $5,000.

```{r echo=FALSE}
ggplot(loans, aes(loan_amount)) +
  geom_histogram(binwidth = 250) +
  coord_cartesian(xlim = c(0,25000))
```

The loan terms in this data set fall into three bins: 12, 36, or 60 months, with 
36 months being the most common. Since there are only 3 discrete values, I am 
going to change the vector type from integer to a factor, so I can graph it
categorically in the bivariate section. I would also find the length-of-loan 
variable easier to interpret as 1, 3, or 5 years.

```{r echo=FALSE}
#factor loan_term_months based on observations in previous graph
loans$loan_term_years <- factor(loans$loan_term_months / 12)
ggplot(loans, aes(loan_term_years)) +
  geom_bar()
```

It looks like borrowers' credit scores are normally distributed, 
with the peak in the 685-695 bin. 

```{r echo=FALSE}
#average the upper and lower credit scores so it's easier to visualize and 
#compare to other variables laater
loans <- mutate(loans, 
    credit_score_average = (credit_score_upper + credit_score_lower)/2)

ggplot(aes(x=credit_score_average), data = loans) +
  geom_histogram(binwidth = 10) +
  scale_x_continuous(limits = c(400, 900))
```

It looks like the most commonly occurring loan statuses are current, completed, 
and charged off. A charged off loan is one where the borrower defaulted and any
remaining loan balance will likely not be collected. 

```{r echo=FALSE}
ggplot(aes(x = loan_status_current), data = loans) + 
  geom_bar() +
  coord_flip()
```

Monthly income appears to be normally distributed around a peak of about $3,000, 
with a long tail out into higher values.

```{r echo=FALSE}
ggplot(aes(x=stated_income_monthly), data = loans) +
  geom_histogram(binwidth = 500) +
  xlim(0,15000)
```

Most borrowers are mployeed. I don't have access to the metadata to understand 
the difference between full-time and employed.

```{r echo=FALSE}
ggplot(loans, aes(employment_status)) +
  geom_bar() +
  coord_flip()
```

Employment duration:

```{r echo=FALSE}
loans <- mutate(loans,
                employment_duration_years = employment_duration_months/ 12)

ggplot(loans, aes(employment_duration_years)) +
  geom_histogram(binwidth = .5) +
  xlim(0,20)
```


```{r echo=FALSE}
summary(loans$employment_duration_years)
```

I created a new variable called lender_profit_current. The formula is principal 
paid plus interest and fees paid minus the original loan amount. 

```{r echo=FALSE}
loans <- mutate(loans,
       lender_profit_current = principal_paid + interest_and_fees_paid - 
       loan_amount)
    
ggplot(aes(x = lender_profit_current), data = loans) +
  geom_histogram(binwidth = 500) + 
  xlim(c(-25000,10000))

summary(loans$lender_profit_current)
``` 


The following graph shows lender_profit_current for only Completed or Chargedoff 
loans. It looks like closed loans are more scewed to positive profit for the 
bank than all loans (open and closed). This makes sense.

```{r echo=FALSE}
#what are distinct factors in loans$loan_status_current?
unique(loans$loan_status_current)

closed_loans <- filter(loans, loan_status_current == 'Completed' | 
                         loan_status_current == 'Chargedoff')

ggplot(aes(x = lender_profit_current), data = closed_loans) +
  geom_histogram(binwidth = 500) + 
  xlim(c(-25000,10000))

summary(closed_loans$lender_profit_current)
```

Mean bank profit on closed loans: Average profit of $21.12 seems low to me.

```{r echo=FALSE}
summary(closed_loans$lender_profit_current)
```

*** Making switch to check out just closed loans -- look at historical trends
pleted loans are loans where the borrower completely paid off the original 
loan amount. 24.0 percent of closed loans are in the Chargedoff category. 

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_status_current)) + geom_bar()
```

### Bivariate Analysis - Charged Off and Completed Loans Only

The spread for closed loans leans more towards smaller loans than the full
Prosper set. Perhaps consumers were asking for smaller loans historically, or
perhaps Prosper was more conservative in how much funding they would award to 
consumers as they refined their profit predictions -- I don't have a way to find
the answer to that question with the dataset that I have. 

Most popular loans: $4,000, $5,000, $3,000, $2,000

With these smaller loan sizes, this smaller profit makes more sense. This data
was also collected over the span of the recession, so interest rates would have
been very low during that time. That also helps explain the low ROI per loan.

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_amount)) +
  geom_histogram(binwidth = 250) +
  coord_cartesian(xlim = c(0,25000))
```

Three-year loans are still the  most common. In relation to the 3-year loans, 
there are proportionally more 1-year loans and less 5-year loans as compared
to the full Prosper loan dataset. 

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_term_years)) +
  geom_bar()
```

Average Credit Score 

```{r}
ggplot(closed_loans, aes(credit_score_average)) +
  geom_histogram(binwidth = 10) +
  xlim(c(400,900))
```


```{r}
ggplot(closed_loans, aes(stated_income_monthly)) +
    geom_histogram(binwidth = 500) +
  xlim(0,15000)

summary(closed_loans$stated_income_monthly)
```

```{r}
ggplot(closed_loans, aes(employment_status)) +
  geom_bar() +
  coord_flip()
```

This graph shows what the bank estimated the return to be when they funded the
loan. It looks like there is a spread between 0.025 and 0.2 (or 2.5 to 20 
percent of return on investment).

```{r echo=FALSE}
closed_loans %>%
  filter(!is.na(estimated_return)) %>%
  ggplot(aes(estimated_return)) +
    geom_histogram(binwidth = .01) +
    coord_cartesian(xlim = c(-0.1,0.25))
```


```{r echo=FALSE}
closed_loans <- mutate(closed_loans,
    actual_return = lender_profit_current / loan_amount)

ggplot(closed_loans, aes(actual_return)) + 
  geom_histogram(binwidth = .025) 
```

Most loans are profitable for the bank, with more common profits ranging from 
just breaking even (.0) to 50 percent (.5). There is a strange peak above 0.5,
and I'm not sure why that is. If I were working for Prosper, I would definitely
look at that. But for the sake of this project, I'm going to move past it. 

The formula I used for acual return was lender_profit_current divided by 
loan_amount. 


```{r echo=FALSE}

#difference <- closed_loans$actual_return - closed_loans$estimated_return

#head(difference, 10)
# graph isn't working -- need to figure out why there are na values in result
# I didn't think I had na values in actual_return
# ggplot(difference, aes(difference)) + geom_histogram()
```


```{r echo=FALSE}
closed_loans %>%
  filter(!is.na(debt_income_ratio)) %>%
  ggplot(aes(debt_income_ratio)) +
    geom_histogram(binwidth = 0.5)
```

And zoomed in to remove that outlier column:

```{r echo=FALSE}
closed_loans %>%
  filter(!is.na(debt_income_ratio)) %>%
  ggplot(aes(debt_income_ratio)) +
    geom_histogram(binwidth = 0.05) +
    xlim(0,1.5)
```

The graph peaks at the 0.125-0.175 bin (or 12.5 to 17.5 percent debt-to-income
ratio).

 
Did some transformations. 
Now, I have the following variables to move into the bivariate analysis: 

```{r echo=FALSE}
# Re-order columns to make more sense logically; drop variables that I am not
# going to look at further.

closed_loans <- select(closed_loans, 
                       loan_status_current, loan_term_years, estimated_return, 
                       loan_amount, principal_paid, interest_and_fees_paid, 
                       lender_profit_current, actual_return, employment_status,
                       employment_duration_years, stated_income_monthly, 
                       credit_score_average, debt_income_ratio)

dim(closed_loans)
```

```{r echo=FALSE}
names(closed_loans)
```


## Bivariate Exploration {#bivariate}

[Back to Top](#top)

Let's start by looking at the correlations between different variables. 
```{r echo=FALSE}
ggcorr(closed_loans, label = TRUE)
```

It looks like there may be a negative correlation between annual income and 
estimated return. My primary factor of interest is estimated_return (displayed
across the bottom). Some of the strong correlations -- between principal paid
and loan_amount or actual_return and lender_profit_current -- are obvious because
the variables do have an explicit relationship. 


```{r echo=FALSE, Bivariate_Plots}
ggplot(closed_loans, aes(credit_score_average,actual_return)) +
  geom_point(alpha = .05) +
  xlim(400,900) +
  geom_hline(yintercept=0, color = 'red', size = 1)
```

It doesn't look like I will be able to use credit score in a predictive model
for actual return. I really expected more of a correlation. It does look like
individuals who have a score over 800 are a very safe bet for investment.
Perhaps the overplotting on this graph is hiding a trend that is there. 


```{r echo=FALSE}
ggplot(closed_loans, aes(credit_score_average, actual_return)) +
  geom_boxplot(aes(group = cut_width(credit_score_average,10))) +
  coord_cartesian(xlim = c(440,880), ylim = c(-.75,1)) +
  geom_hline(yintercept = 0, color='red', size=1) 
```

It looks like the median return is above 0 for all ranges except those
centering at 450 and below. And the peak in the median return occurs at a 
credit score of 570. 

With this graph, it looks like there is a pattern in the relationship between
credit score and return. Those with poor or fair credit scores are bringing the
highest rate of return. This is likely because these individuals pay higher 
interest rates. 

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_term_years,actual_return)) +
  geom_boxplot()
```

It looks like three-year loans have the highest median percentage of return. And
it looks like Prosper is losing the most money on five-year loans. In fact, 
almost half of five-year loans are losing money.

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_amount, actual_return)) +
  geom_point(alpha = .05) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

Here we see the effects of overplotting within the bands of most common loan
amounts. It does look like the return falls much more above 0, which is good
for Prosper. Let's take a look at this with a boxplot grouping to see where 
the medians lie by $1,000 bins.

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_amount,actual_return)) +
  geom_boxplot(aes(group = cut_width(loan_amount, 1000))) +
  xlim(0,25500) +
  coord_cartesian(ylim = c(-.5,.5)) +
geom_hline(yintercept=0, color = 'red', size = .75)
```

Return is above 0 for all loan amounts. There really is no evidence of variation
in the loan amount changing the ROI. 


```{r echo=FALSE}
ggplot(closed_loans,aes(debt_income_ratio,actual_return)) +
  geom_point(alpha = .1) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

The relationship between bank profit and stated monthly income doesn't seem
to have a linear pattern, but an outlier is seriously skewing the x scale of 
that graph, so I'm going to zoom in.

```{r echo=FALSE}
ggplot(closed_loans,aes(debt_income_ratio,actual_return)) +
  geom_point(alpha = .1) +
  coord_cartesian(xlim=c(-.25,2.5)) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

It seems as though a higher debt_income_ratio (DIR) is related to a higher 
profit. There seem to be some tiers to the data, althought it is still hard to see 
due to overplotting. But it seems that there are jumps in the level of profit at
both 0.25 and 0.5 marks for DIR.

```{r echo=FALSE}
ggplot(closed_loans, aes(reorder(employment_status,actual_return, FUN = median), 
  actual_return)) +
  geom_boxplot() +
  coord_flip() 
```

These are some strange tiers, with redundancy. My guess is that the option set 
changed at some point in data collection. For example "employed" was later 
changed to two options -- "part-time" and "full-time."

```{r echo=FALSE}
closed_loans %>%
  #ignore fields that are: blank, 'Not available,' and 'Other'
  filter(employment_status == 'Employed' | 
         employment_status == 'Full-time' |
         employment_status == 'Not employed' |
         employment_status == 'Part-time' |
         employment_status == 'Retired' |
         employment_status == 'Self-employed'   ) %>%
  ggplot(aes(actual_return, y = ..density..)) +
    geom_freqpoly(aes(color=employment_status),binwidth=.025)
```

The plot shows density, so that the count of each category doesn't scew the 
graph. The graph shows a proportion for each category.

It is surprising to me that those with an employment_status of 'employed' bring
Prosper a lower actual_return. However, I know that some of the data collection
criteria changed in 2009, based on the data dictionary. I don't know exactly
when the employment_status optionset would have changed. But if 'Employed' is 
from an earlier date, it's not a good comparison (because there are many 
environmental differences that could make a difference if the data was collected
during two different time periods).

Let's zoom in to the peak area of this graph.

```{r echo=FALSE}
closed_loans %>%
  #ignore fields that are: blank, 'Not available,' and 'Other'
  filter(employment_status == 'Employed' | 
         employment_status == 'Full-time' |
         employment_status == 'Not employed' |
         employment_status == 'Part-time' |
         employment_status == 'Retired' |
         employment_status == 'Self-employed'   ) %>%
  ggplot(aes(actual_return, y = ..density..)) +
    geom_freqpoly(aes(color=employment_status),binwidth=.025) +
    xlim(c(0,0.5))
```

It's interesting that retired and part-time individuals seem to peak at a higher
actual_return rate. Perhaps it is a similar phenomenon to why Prosper makes a 
better return for those with lower credit scores -- perhaps these individuals 
are assigned a higher interest rate. 

```{r echo=FALSE}
ggplot(closed_loans, aes(employment_duration_years,actual_return)) +
  geom_point(alpha = .1) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

```{r echo=FALSE}
ggplot(closed_loans, aes(employment_duration_years,actual_return)) +
  geom_boxplot(aes(group = cut_width(employment_duration_years,1))) +
  xlim(0,20) +
  ylim(-.5,.5) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

```{r echo=FALSE}
ggplot(closed_loans, aes(stated_income_monthly,actual_return)) +
  geom_point(alpha = .1) +
  xlim(0,200000) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```
 
```{r echo=FALSE}
ggplot(closed_loans, aes(stated_income_monthly,actual_return)) +
  geom_boxplot(aes(group = cut_width(stated_income_monthly,10000))) +
  xlim(0,200000) +
  ylim(-.5,.5) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

It looks like debt-to-income ratio is higher in the middle regions of the 
credit score range. I believe that debt to income ratio is one factor that
influences credit score, so it makes sense that those with an excellent credit
score would have a low debt to income ratio. I wonder if those with a score 
under 530 simply have a harder time getting a loan and therefore don't have as 
much debt.

```{r echo=FALSE}
ggplot(closed_loans, aes(factor(credit_score_average), debt_income_ratio)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0,1)) +
  theme(axis.text.x = element_text(angle = 90))

```

* Is a longer employment history correlated with a better credit score?
A longer employment duration does not appear to correlate with a better 
credit score.

```{r echo=FALSE}
closed_loans %>%
  filter(employment_status == 'Employed' | employment_status == 'Full-time' |
           employment_status == 'Part-time' | 
           employment_status == 'Self-employed') %>%
  ggplot(aes(credit_score_average, employment_duration_years)) +
    geom_point(alpha = .1) +
    xlim(450,850)
```

### Bivariate Summary

Loan term covaried with actual return. The return for one-year loans was mostly
positive, but not much more than 0. Three-year loans brought in the highest 
return. Five-year loans were responsible for the most lost in return. 

Credit score also varied with actual return. Credit scores centering right under 
600 actually brought the highest level of return. 

Otherwise, I really didn't find any other patterns of covariation. 

## Multivariate Exploration {#multivariate}

[Back to Top](#top)

Question:

* Does loan term in years add any meaningful pattern to a graph of actual_return
and credit_score_average?

```{r echo=FALSE, Multivariate_Plots}
ggplot(closed_loans, aes(factor(credit_score_average), actual_return, 
                         fill=factor(loan_term_years))) +
  geom_boxplot() +
  scale_fill_brewer(type='seq', guide=guide_legend(title='Loan Term')) +
  theme(axis.text.x = element_text(angle = 90))
```

Within loans that are three years in duration (the most commonly occurring and
easiest to see), it looks like the median return is highest in the upper 500s 
and lower 600s. The median return steadily increases above a credit score of 
630.

```{r, echo=FALSE}
ggplot(closed_loans, aes(credit_score_average,actual_return,
                         color=employment_status)) +
  geom_point(alpha = .5) +
  xlim(400,900) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

In this graph, we can see that the colors associated with 'not available,' 'not
employed,' and 'other' occur more frequently in the lower credit scores, and 
that employment of some type or retirement occurs more frequently in the higher 
credit score ranges. 

```{r echo=FALSE}
ggplot(closed_loans,aes(debt_income_ratio,actual_return,
                        color=employment_status)) +
  geom_point(alpha = .5) +
  coord_cartesian(xlim=c(-.25,2.5)) + stat_ellipse()
```

Adding employment status to this graph doesn't show any new insights. Debt-to-
income ratio combined with empoyment status does not seem to build a pattern.

------

## Summary {#summary}

[Back to Top](#top)

This plot shows that most loans that have a positive return on investment. The 
average return for a funded loan is 2.8 percent.

```{r echo=FALSE, Plot_One}
ggplot(closed_loans, aes(actual_return)) + 
  geom_histogram(binwidth = .025) +
  labs(title = 'Return on Investment for Loan Funding', x='Actual Return*', 
       caption='*Actual return is profit as a percentage of initial loan 
       amount.', y='Count')
```

```{r echo=FALSE}
summarize(closed_loans,
          mean_return_percent = mean(actual_return)*100,
          median_return_percent = median(actual_return)*100)
```

Prosper gives out loans that are one, three, or five years in duration. Almost 
90 percent of loans are provided with a three year term. The median return 
for these loans is 13.6 percent. However, the average return is only 4 percent.
Negative outliers are bringing the average lower than the median. 

The median return for one-year loans is 5.9 percent, with average profit at 3.8
percent. And the median return for five-year loans is 5.8 percent, with average
profit at negative 13.4 percent. As can be seen in the graph above, the second
quartile for five-year loans falls into a negative return. The five-year loan
category is not profitable for Prosper. 

```{r echo=FALSE, Plot_Two}
ggplot(closed_loans, aes(loan_term_years, actual_return)) +
  geom_boxplot() + 
  labs(title ='Actual Return by Loan Duration', x='Loan Term (Years)', 
       y='Actual Return*', caption = '*Actual return is profit as a percentage 
       of initial loan amount.')
```


```{r echo=FALSE}
closed_loans %>%
  group_by(loan_term_years) %>%
  summarize(
    count = n(),
    return_mean_percent = mean(actual_return, na.rm = TRUE)*100,
    return_median_percent = median(actual_return, na.rm = TRUE)*100
  )
```

This graph shows that the median profit is above 0 for almost all credit score
categories except those below 470. However, if we look at the mean profit, 
return becomes reliable above a credit score of 530.

```{r echo=FALSE, Plot_Three}
ggplot(closed_loans, aes(credit_score_average, actual_return)) +
  geom_boxplot(aes(group = cut_width(credit_score_average,20))) +
  coord_cartesian(xlim = c(440,880), ylim = c(-1,1.25)) +
  geom_hline(yintercept=0, color = 'red', size = .75) + 
  labs(title ='Actual Return by Borrower Credit Score', x='Credit Score', 
    y='Actual Return*', caption = '*Actual return is profit as a percentage of 
    initial loan amount.')
```

```{r echo=FALSE}
closed_loans %>%
  group_by(cut_width(credit_score_average,20)) %>%
  summarize(
    count = n(),
    return_mean_percent = mean(actual_return, na.rm = TRUE)*100,
    return_median_percent = median(actual_return, na.rm = TRUE)*100
  ) %>% rename("bins" = "cut_width(credit_score_average, 20)")

```


### Further areas for investigation: 

* How does early payoff impact actual return? That is, if a borrower pays their
loan off early, I anticipate that actual return would be lower.  
* Analyze the difference between estimated return and actual return. Focus on 
areas where the bank anticipated profit but lost money. Is there a factor that
their predictions should also take into account? Or is that just part of the 
risk of the lending process (e.g. the exploration would not uncover additional 
factors to create better predictive models of expected profit)? 
